Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-30 parallel-tractability-J-v9/response.tex

\begin{quote}
There should, however, be some discussion somewhere in the paper on the
relationship between path length and KB size.   Even if a KB uses a
non-parallelizable ontology language this doesn't matter much if the paths
that cause problems are short compared to the size of the KB.  The
non-parallel aspects of path processing of short paths will not be noticable
with a small number of threads (certainly up to hundreds) and maybe not even
for very many threads (because of general overhead).
\end{quote}

We added xxx to adress this...

\begin{quote}
  The second sentence of Section 3 is badly stated and misleading.  It is definitely true that for a Datalog program $\langle R, I \rangle$ the rule set $R$ is fixed.  However, the initial facts $I$ are also fixed.
\end{quote}

We have rephrased the beginning of Section~3 and made it clear that the rule set is fixed for a class of Datalog programs.
We also revised the similar presentations in Section~2.

@@

The paper has so many grammatical errors near its beginning that I gave up
marking them after the first two pages.  
\end{quote}

TODO: Remove the quote above once the issues are fixed.

The paper has been carefully proof-read and hopefully all wording and structural issues are fixed now.

\begin{quote}
"parallely tractable" is very grating.  It would be better to rewrite
sentences to not use it.
\end{quote}

TODO: once done, check whether indeed all occurrences are removed

We have removed all occurrences of ``parallely tractable'' and
rephrased the sentenced to use ``tractable in parallel'' or ``parallel
tractability'' instead.

### 2018-01-30 parallel-tractability-J-v9/parallelTractabilityDatalog.tex
Section 3 up to optimized algo proof read

%\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
 \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
\usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm}

\usepackage{amsmath}

\usepackage{CJK}
\usepackage{color}

\usepackage{enumitem}
\usepackage{rotating}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

%\journal{Nuclear Physics B}

\hyphenation{data-log}

\begin{document}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{Parallel Tractability of Ontology\\
Materialization: Technique and Practice}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \address[label1]{}
%% \address[label2]{}

\author{}

\address{}

\begin{abstract}
Materialization is an important reasoning service for applications built
on the ontology languages, such as OWL. With a
rapid growth of semantic data, it is challenging to perform
materialization on large-scale ontologies efficiently.
Current work proposes parallel algorithms and employs computing
platforms of high-performance to make materialization sufficiently efficient and scalable in practice.
However, there do exist ontologies, for which
materialization cannot always be improved by parallelism (we use experiments to show it in Section~\ref{sec:evaluation}).
On the other hand, some well-known large-scale ontologies, such as YAGO, have been shown to have
good performance for parallel reasoning, but they are expressed in ontology languages
that are not parallelly tractable in theory, i.e., the efficiency of reasoning may not be improved
on a parallel implementation.
This motivates us to study the problem of \emph{parallel tractability} of ontology
materialization from a theoretical perspective.
That is, we aim to identify the ontologies for which materialization is
parallelly tractable, i.e., in \texttt{NC} complexity.
The results of the parallel tractability can be further used to optimize
materialization algorithms and to guide ontology engineers in creating parallelly tractable ontologies.

In this work, we focus on datalog rewritable ontology languages.
We propose some algorithms, called \texttt{NC} algorithms, to identify several classes of datalog rewritable
ontologies (called \emph{parallelly tractable classes})
such that materialization over them is parallelly tractable.
We then investigate the parallel tractability of materialization over
two datalog rewritable ontology languages DL-Lite and DHL (\emph{Description Horn Logic}).
We show that materialization over any ontology expressed in DL-Lite$_{core}$ or DL-Lite$_\mathcal{R}$ is parallelly tractable.
For DHL, there exist ontologies where materialization can hardly be parallelized.
We propose to restrict the usage of DHL such that the restricted ontologies are parallelly tractable.
We further extend the results to an extension of DHL that also allows \emph{complex role inclusion axioms}.
To verify the practical usability of the above results, we analyze real-world datasets and show
that many ontologies expressed in DHL or its extension belong to
the parallelly tractable classes. Based on the results of parallel tractability,
we further implement an optimized parallel algorithm
and compare it to the state-of-the-art reasoner RDFox. The experimental results show that the
optimization used in our implementation results in a better
performance on parallelly tractable ontologies compared with RDFox.
\end{abstract}

\begin{keyword}
ontology \sep materialization \sep datalog \sep parallel tractability \sep \texttt{NC} complexity
\end{keyword}

\end{frontmatter}

%% \linenumbers

\section{Introduction}
\label{sec:introduction}

The Web Ontology Language (OWL)\footnote{The latest version is OWL 2: http://www.w3.org/TR/owl2-overview/}
is an important standard for ontology language in the Semantic Web and knowledge-based applications.
In these applications, \emph{materialization}, which is the reasoning task of computing all implicit
facts that follow from a given ontology \cite{handbook}, plays an important role. By means of it, developers and users
can optimize query answering, ontology diagnosis and debugging. Since huge amounts of semantic data
are being generated at an increasing pace by sensor networks, government authorities and social
media \cite{LehmbergRMB16,MeuselBP15}, it is challenging to conduct materialization on such large-scale ontologies efficiently.

There has been work that proposes parallel reasoning algorithms and employs parallel computing platforms 
for the ontology language RDFS and its extended
fragments \cite{MotikNPHO14,PetersSZ15,SubercazeGCL16}. Several optimization strategies, e.g., dictionary encoding, balance workload,
data partitioning, are further studied to enhance parallel RDFS reasoning. There are also parallel implementations for scalable
reasoning of highly expressive ontology languages \cite{SteigmillerLG14,WuH12}. On the other hand, in order to make reasoning tasks more efficient
in parallel, several work utilizes different kinds of high-performance computing platforms, like supercomputers \cite{Hoeksema2011,GoodmanJMAAH11},
MapReduce \cite{UrbaniKMHB12} and GPU servers \cite{HeinoP12}.

The above work verifies the effectiveness of parallel reasoning based on the evaluations on different test datasets.
However, for most popular ontology languages, even RDFS and datalog rewritable ontology languages,
which have \texttt{PTime}-complete or higher complexity of reasoning in the worst case,
they are not parallelly tractable according to \cite{Raymond95}, i.e., the efficiency of reasoning may not be
improved on a parallel implementation. A possible reason of the high performance of parallel
reasoning is that the utilized test datasets do not fall in the worst cases in terms of computational complexity.
It has also been shown that, some well-known large-scale ontologies, such as YAGO, have good performance for parallel
reasoning \cite{KolovskiWE10}, but they are expressed in ontology languages that are not parallelly tractable in theory.
On the other hand, we show in Section~\ref{sec:evaluation} that, there do exist some ontologies, for which materialization cannot
always be improved by parallelism. The current work of parallel ontology reasoning can hardly explain this.
While one can try out different parallel implementations to see whether an ontology can be handled by (one of) them efficiently,
we study the problem of parallel tractability in theory and identify properties that make an ontology parallelly tractable.
These properties can be further used to optimize parallel algorithms and to guide ontology engineers in creating ontologies
for which parallel tractability can be guaranteed theoretically.

According to \cite{MotikNPHO14}, many real large-scale ontologies are essentially expressed in the ontology languages
that can be rewritten into datalog rules. Thus, we focus on such datalog rewritable ontology languages in this paper.
The main target of this paper is to identify the classes of datalog rewritable ontologies such that materialization
over these ontologies is parallelly tractable, i.e., in the parallel complexity class $\texttt{NC}$ \cite{Raymond95}.
This complexity class consists of problems that can be solved efficiently in parallel.
To show that a problem is in the \texttt{NC} class, one can give an \emph{\texttt{NC} algorithm} that handles this problem
in parallel computation \cite{Raymond95}.
An \texttt{NC} algorithm is required to terminate in parallel poly-logarithmic time.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not \texttt{NC} algorithms,
since they are designed for general datalog programs and has \texttt{PTime}-complete complexity.
Thus, we first give \texttt{NC} algorithms that perform materialization,
and then identify the corresponding classes of datalog rewritable ontologies (called \emph{parallelly tractable classes})
that can be handled by these \texttt{NC} algorithms. To make the proposed \texttt{NC} algorithms practical,
we also discuss how to optimize and implement them.

In the practical part of this work, we study specific datalog rewritable ontology languages.
We first focus on the ontology language DL-Lite to clarify how to apply the above \texttt{NC} algorithms
to study the parallel tractability of ontology materialization.
We show that DL-Lite$_{core}$ and DL-Lite$_\mathcal{R}$ are parallelly tractable.
We then study the ontology language Description Horn Logic (DHL) \cite{GrosofHVD03}, which is the intersection of datalog and OWL
in terms of expressivity. We give a case of a DHL ontology where materialization can hardly be parallelized.
Based on the analysis of this case, we propose to restrict the usage of DHL such that materialization over the
restricted ontologies can be handled by the proposed \texttt{NC} algorithms.
We further extend the results to an extension of DHL that also allows complex role inclusion axioms.
Finally, we analyze well-known benchmarks and real-world datasets and show that many real ontologies following the
proposed restrictions belong to the parallelly tractable classes.
We implement a system based on an optimized \texttt{NC} algorithm for DHL materialization.
We compare our system to the state-of-the-art reasoner RDFox.
The experimental results show that the optimization proposed in this paper results in a better performance on parallelly
tractable ontologies compared with RDFox.

The rest of the paper is organized as follows. In Section~\ref{sec:background}, we introduce some basic notions.
We then give two \texttt{NC} algorithms for ontology materialization in Section~\ref{sec:ptclass}.
We study the parallelly tractable materialization of DL-Lite, DHL and an extension of DHL in Section~\ref{sec:ptonto} respectively.
In Section~\ref{sec:practicalAlg},
we discuss how to optimize and implement the given \texttt{NC} algorithms.
We analyze real-world datasets and evaluate our implementation in Section~\ref{sec:evaluation}.
We then discuss related work in Section~\ref{sec:related} and conclude in Section~\ref{sec:conclusion}.
The detailed proofs of the theorems and the lemmas in this paper are given in the appendix.

\section{Background Knowledge}
\label{sec:background}

In this section, we introduce some basic notions that are needed to introduce our approach.

\subsection{Datalog}

We discuss the main issues in this paper using standard datalog notions.
In datalog \cite{database}, a \emph{term} is a variable or a constant. An \emph{atom} $A$
is defined by $A\equiv p(t_1,...,t_n)$ where $p$ is a \emph{predicate} (or \emph{relational})
name, $t_1,...,t_n$ are terms, and $n$ is the arity of $p$. If all the terms in an atom $A$ are
constants, then $A$ is called a \emph{ground atom}.
A datalog \emph{rule} is of the form: `$B_1,...,B_n\rightarrow H$',\footnote{In datalog rules, a comma
represents a Boolean conjunction `$\wedge$'.} where $H$ is referred to as
the \emph{head atom} and $B_1,...,B_n$ the \emph{body atoms}. Each variable in the head atom
of a rule must occur in at least one body atom of the same rule. A \emph{fact} is a rule of
the form `$\rightarrow H$', i.e., a rule with an empty body and the head $H$ being a ground atom.
A datalog program $P$ consists of rules and facts.
A \emph{substitution} $\theta$ is a partial mapping of variables to constants.
For an atom $A$, $A\theta$ is the result of replacing each variable $x$ in $A$
with $\theta(x)$ if the latter is defined. We call $\theta$ a \emph{ground substitution}
if each defined $A\theta$ is a ground atom.
A \emph{ground instantiation} of a rule is obtained by applying a ground substitution on all
the terms in this rule with respect to a finite set of constants occurring in $P$.
Furthermore the ground instantiation of $P$, denoted by $P^*$,
consists of all ground instantiations of rules in $P$.
The predicates occurring only in the body  of some rules are called \emph{EDB predicates},
while the predicates that may occur as head atoms are called \emph{IDB predicates}.

\subsection{DHL and DL-Lite}

In what follows, we use $\textbf{CN}$, $\textbf{RN}$ and $\textbf{IN}$
to denote three disjoint countably
infinite sets of \emph{concept names, role names}, and \emph{individual names} respectively.
The set of roles is defined as $\textbf{R}:=\textbf{RN}\cup\{R^-|R\in\textbf{RN}\}$
where $R^-$ is the \emph{inverse role} of $R$.
For ease of discussion, we focus on the \emph{simple forms} of axioms shown
in the left column of Table~1. These simple forms can be obtained by using
well-known \emph{structure transformation} techniques \cite{KrotzschRH07,Kazakov09}.

DHL (short for \emph{description horn logic}) \cite{GrosofHVD03} is introduced as an
intersection of description logic (DL) and datalog in terms of expressivity.
We define a DHL ontology $\mathcal{O}$ as a triple:
$\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$, where
$\mathcal{T}$ denotes the TBox containing axioms of the forms (T1-T3);
$\mathcal{R}$ is the RBox that is a set of axioms of the forms (R1-R3);
$\mathcal{A}$ is the ABox containing \emph{assertions} of the forms (A1) and (A2).
In an axiom of either of  the forms (T1-T3 and R1-R3), concepts $A_{(i)}$ and $B$ are either
concept names, \emph{top concept} ($\top$) or \emph{bottom concept} ($\bot$); $R$ and $S_{(i)}$
are roles in $\textbf{R}$.
For an axiom of the form $A\sqsubseteq\forall R.B$ that is also allowed in DHL, we only consider its
equivalent form $\exists R^-.A\sqsubseteq B$.

DHL is related with other ontology languages.
First, DHL is essentially a fragment of the description logic Horn-$\mathcal{SHOIQ}$ with
disallowing \emph{nominal}, \emph{number restriction} and
right-hand \emph{existential restriction} ($A\sqsubseteq\exists R.B$).
Second, the expressivity of DHL covers that of RDFS to some extent \cite{GrosofHVD03}.
Reasoning with RDFS ontologies is \texttt{NP}-complete \cite{Horst05}
and, thus, is not parallelly tractable.
However, by applying some simplifications and restrictions, RDFS ontologies can be
expressed in DHL \cite{GrosofHVD03} that has \texttt{PTime}-complete complexity for
materialization.

\begin{table}
\begin{center}
\begin{tabular}{lcc}
\multicolumn{3}{c}{\textbf{Table 1: Axioms and corresponding datalog rules}}\\
\hline
&~~~~~~~~~~~~~~~~~~~~~~~~Axioms~~~~~~~~~~~~~~~~~~~~~~~~&~~~~~~~~~~~~~~~~~~~~~~~~Datalog Rules~~~~~~~~~~~~~~~~~~~~~~~~\\
\hline
\hline
(T1)& $A\sqsubseteq B$& $A(x)\rightarrow B(x)$\\
(T2)& $A_1\sqcap A_2\sqsubseteq B$& $A_1(x), A_2(x)\rightarrow B(x)$\\
(T3)& $\exists R.A\sqsubseteq B$& $R(x,y), A(y)\rightarrow B(x)$\\
(T4)& $A\sqsubseteq\exists R$& $A(x)\rightarrow R(x, o_{R}^A)$\\

\hline
(R1)& $S\sqsubseteq R$& $S(x,y)\rightarrow R(x,y)$\\
(R2)& $S\sqsubseteq R^-$& $S(x,y)\rightarrow R(y,x)$\\
(R3)& $R\circ R\sqsubseteq R$& $R(x,y), R(y,z)\rightarrow R(x,z)$\\
(R4)& $R_1\circ R_2\sqsubseteq R$&$R_1(x,y),R_2(y,z)\rightarrow R(x,z)$\\

\hline
(A1)& $A(a)$& $A(a)$\\
(A2)& $R(a,b)$& $R(a,b)$\\
\hline
\end{tabular}
\label{tab:dhl}
\end{center}
\end{table}

In the initial work of DHL \cite{GrosofHVD03}, \emph{complex role inclusion axioms} (complex RIAs) of
the form $R_1\circ...\circ R_n\sqsubseteq R$ are not considered, although they can be
naturally transformed to datalog rules.
In this paper, we also consider an extension of DHL (denoted by DHL($\circ$))
that allows complex RIAs. Since a complex RIA can be transformed to
several axioms of the form (R4), we then require that an
RBox $\mathcal{R}$ of a DHL($\circ$) ontology can contain
axioms of the forms (R1-R4). Note that (R3) is actually a special
case of (R4).

DL-Lite is a group of ontology languages designed for highly-efficient
query answering through knowledge
and underpins the OWL profile OWL QL \cite{CalvaneseGLLR07}.
DL-Lite$_{core}$ and DL-Lite$_{\mathcal{R}}$ are the two basic fragments
in DL-Lite.
DL-Lite$_{core}$ requires that an TBox contains axioms of the forms (T1) $A\sqsubseteq B$,
(T2) $A_1\sqcap A_2\sqsubseteq B$ where $B\equiv\bot$, (T3) $\exists R.A\sqsubseteq B$
where $A\equiv\top$ or $B\equiv\bot$, and (T4) $A\sqsubseteq\exists R$,
an ABox contains \emph{assertions} of the forms (A1) and (A2), and
RBox is not involved. DL-Lite$_{\mathcal{R}}$ is an extension of DL-Lite$_{core}$,
which allows involving RBoxes that contain axioms of the forms (R1-R2).
If not specially specified, \emph{a DL-Lite ontology denotes a DL-Lite$_{\mathcal{R}}$ ontology
in the following paragraphs}.

\subsection{Ontology Materialization via Datalog Programs}

An ontology
expressed in DHL, DHL($\circ$), DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ can be transformed to a datalog program
(see the corresponding rules in the right column of Table~1).
Note that, for an axiom of the form (T4) $A\sqsubseteq\exists R$, it should be transformed to
a first-order logic rule `$A(x)\rightarrow \exists y(R(x,y))$' where $y$ is a free variable.
In this work, such a free variable $y$ is eliminated via Skolemisation into a new individual $o_{R}^A$
that corresponds to concept $A$ and role $R$. One can check that the elimination of free variables
does not influence the results of materialization by according to \cite{CalvaneseGLLR07}.

In what follows, for an ontology $\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$,
we also use $P=\langle R, \textbf{I}\rangle$ to represent the corresponding datalog program
where $R$ is the set of rules transformed from the axioms in $\mathcal{T}$ and $\mathcal{R}$,
$\textbf{I}$ is the set of facts that are directly copied from the assertions in $\mathcal{A}$.
Further, we use $R_1\sqsubseteq_{*}R_2$ to denote the smallest transitive reflexive relation
between roles such that $R_1\sqsubseteq R_2\in\mathcal{R}$ implies $R_1\sqsubseteq_{*}R_2$
and $R_1^-\sqsubseteq_{*}R_2^-$. In this paper, we also use the
notion of \emph{simple role}, which is initially proposed to restrict the
usage of highly expressive ontology languages \cite{HorrocksS04}.
Specifically, a role $S\in\textbf{R}$ is \emph{simple} if, (1) it has no subrole (including $S$)
occurring on the right-hand side of axioms of the forms (R3) and (R4); (2) $S^-$ is simple.

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
let $T_R(\textbf{I})=\{H\theta|\forall B_1,...,B_n\rightarrow H\in R, B_i\theta\in\textbf{I} (1\leq i\leq n)\}$,
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}.

\subsection{The $\texttt{NC}$ Complexity}

The parallel complexity class $\texttt{NC}$, known as Nick$'$s
Class \cite{Raymond95}, is studied by theorists as a parallel complexity class
where each decision problem can be efficiently solved in parallel.
Specifically, a decision problem in the $\texttt{NC}$ class
can be solved in poly-logarithmic time on a parallel machine with a polynomial number of processors.
We also say that an $\texttt{NC}$ problem can be solved in \emph{parallel poly-logarithmic time}.
Although the $\texttt{NC}$ complexity is a theoretical analysis tool,
it has been shown that many $\texttt{NC}$ problems can be solved efficiently in practice \cite{Raymond95}.

From the perspective of implementations, the $\texttt{NC}$ problems are also highly
parallel feasible for other parallel models like BSP \cite{Valiant90}
and MapReduce \cite{KarloffSV10}. The \texttt{NC} complexity is originally defined
as a class of decision problems. Since we study the problem of materialization, we do not
require in this work that a problem should be a decision problem in $\texttt{NC}$.
In addition, since many parallel reasoning systems (see related work in Section~\ref{sec:related})
are implemented on shared-memory platforms, we
study all the issues in this work by assuming that the running machines are in
shared-memory configurations.

@@

Our target is to find for which kinds of ontologies (not ontology languages)
materialization is parallelly tractable.
Since we assume that for any datalog program $\langle R, \textbf{I}\rangle$ the rule
set $R$ is fixed, the materialization problem is thus in data complexity
$\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
case \cite{Raymond95}. In other words, the materialization problem on general
datalog programs cannot be solved in parallel poly-logarithmic time unless \texttt{P}=\texttt{NC}.
Thus, we say that \emph{materialization on a class of datalog programs is parallelly tractable
if there exists an algorithm that handles this class of datalog programs and runs in parallel
poly-logarithmic time (this algorithm is also called an \texttt{NC} algorithm)}.
In this section, we identify such classes of datalog programs by studying different \texttt{NC} algorithms.

\subsection{Parallel Tractable Classes}

We first give the following definition for a parallel tractable class of datalog programs
where an \texttt{NC} algorithm exists for handling each datalog program in this class.

\begin{definition}\label{def:ptd}
\textbf{(Parallelly Tractable Class)} Given a class $\mathcal{D}$ of datalog programs,
we say that $\mathcal{D}$ is a parallelly tractable datalog program (\texttt{PTD}) class
if there exists an \texttt{NC} algorithm that performs sound and complete materialization for each datalog program
in $\mathcal{D}$. The corresponding class of ontologies of $\mathcal{D}$ is called a
parallelly tractable ontology (\texttt{PTO}) class.
\end{definition}

According to the above definition, if we find an \texttt{NC} algorithm $\mathsf{A}$
for datalog materialization, then we can identify a \texttt{PTD} class $\mathcal{D}_{\mathsf{A}}$,
which is the class of all datalog programs that can be handled by $\mathsf{A}$.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not \texttt{NC} algorithms
due to their \texttt{PTime}-complete complexity, since they are designed for handling general datalog programs.
Thus we give our $\texttt{NC}$ algorithms.
In the following, we first give a parallel materialization algorithm that works for
general datalog programs. We then restrict this algorithm to an \texttt{NC} version
and identify the target \texttt{PTD} class.

@@

In order to give a parallel materialization algorithm,
we introduce the notion of \emph{materialization graph}.
It makes the analysis of the given algorithm convenient.

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$;
\end{enumerate}
~~$\mathcal{G}$ satisfies the following condition:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $\forall H,B_1,...,B_n\in V$ such that $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$, we have that
  $B_1,...,B_n\rightarrow H\in P^*$.
\end{enumerate}
\end{definition}

For some derived atom $H$, there may exist several rule instantiations where $H$
occurs as a head atom.
This also means that $H$ can be derived in different ways.
The condition in the definition above results in only one way of deriving $H$ being
described by a materialization graph.
Suppose $\mathcal{G}$ is a materialization graph, the nodes whose in-degree is 0 are the
original facts in $\textbf{I}$.
The size of $\mathcal{G}$, denoted by $|\mathcal{G}|$, is the number of nodes in $\mathcal{G}$.
The depth of $\mathcal{G}$, denoted by \texttt{depth}($\mathcal{G}$), is the maximal length of a path
in $\mathcal{G}$.
We next give an example of a materialization graph.\\

\begin{example}\label{exp:mg}
Consider a DHL($\circ$) ontology $\mathcal{O}_{ex_1}$ where the TBox is
$\{\exists R.A\sqsubseteq A\}$, the RBox is $\{S\circ R\sqsubseteq R\}$ and
the ABox is $\{A(b),R(a_1,b),S(a_i,a_{i-1})\}$ for $2\leq i\leq k$ and
$k$ is an integer greater than $2$.
The corresponding datalog program of this ontology is $P_{ex_1}=\langle R, \textbf{I}\rangle$
where $\textbf{I}$ contains all the assertions in the ABox
and $R$ contains the two rules `$R(x,y),A(y)\rightarrow A(x)$'
and `$S(x,y),R(y,z)\rightarrow R(x,z)$'.
The graph in Figure~\ref{fig:mg} is a materialization graph with respect to $P_{ex_1}$,
denoted by $\mathcal{G}_{ex_1}$.
The nodes whose in-degree is 0 are the original facts in $\textbf{I}$;
each of the other nodes corresponds to a ground instantiation of some rule.
For example, the node $A(a_k)$ corresponds to the ground rule instantiation
`$R(a_k,b),A(b)\rightarrow A(a_k)$'.
The size of this materialization graph is the number of nodes, that is $3k$.
The depth of $\mathcal{G}_{ex_1}$ is $k$.
\end{example}

@@

We say that a materialization graph $\mathcal{G}$ is a \emph{complete materialization graph}
when $\mathcal{G}$ contains all ground atoms in $T_R^{\omega}(\textbf{I})$.
The set of nodes in a complete materialization graph is actually the result
of materialization.
Thus, the procedure of materialization can be transformed to the construction of
a complete materialization graph.
We pay our attention to complete materialization graphs and do not distinguish it to the
notion `materialization graph'.
It should also be noted that there may exist several materialization graphs for a datalog program.

\subsection{A Basic Parallel Algorithm}

>>>>

@@

Our target is to find for which kinds of ontologies (not ontology languages)
materialization is tractable in parallel. We consider data complexity,
i.e., we consider classes of datalog programs which share a fixed the rule set. 
The data complexity of the materialization problem in this case is
$\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
case \cite{Raymond95}.
% Since we assume that, for any class of datalog programs where the rule
% set of each datalog program is fixed, the materialization problem is, thus, in data complexity
% $\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
% case \cite{Raymond95}.
In other words, the materialization problem of 
datalog programs cannot be solved in parallel poly-logarithmic time unless \texttt{P}=\texttt{NC}.
Thus, we say that \emph{materialization for a class of datalog programs is tractable in parallel
if there exists an algorithm that handles this class of datalog programs and runs in parallel
poly-logarithmic time. Such an algorithm is also called an \texttt{NC} algorithm.}
In this section, we identify such classes of datalog programs by studying different \texttt{NC} algorithms.

\subsection{Parallel Tractability Classes}

We first give the following definition for a class of datalog programs
that is tractable in parallel, i.e., an \texttt{NC} algorithm exists for handling each datalog program in this class.

\begin{definition}\label{def:ptd}
\textbf{(Parallel Tractability Class)} Given a class $\mathcal{D}$ of
datalog programs sharing the same rule set, we say that $\mathcal{D}$
is a class of datalog program with \emph{p}arallel
\emph{t}ractability, short a \emph{\texttt{PTD} class}, 
if there exists an \texttt{NC} algorithm that performs sound and complete materialization for each datalog program
in $\mathcal{D}$. The corresponding class of ontologies of $\mathcal{D}$ is called a
class of ontologies with parallel tractability, short a \texttt{PTO} class.
\end{definition}

According to the above definition, if we find an \texttt{NC} algorithm $\mathsf{A}$
for datalog materialization, then we can identify a \texttt{PTD} class $\mathcal{D}_{\mathsf{A}}$,
which is the class of all datalog programs that can be handled by $\mathsf{A}$.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not \texttt{NC} algorithms
due to their \texttt{PTime}-complete complexity, since they are
designed for handling general datalog programs. Thus, we proceed to
devise specific \texttt{NC} algorithms.
In the following, we first give a parallel materialization algorithm that works for
general datalog programs. We then restrict this algorithm to an \texttt{NC} version
and identify the target \texttt{PTD} class.

@@

In order to give a parallel materialization algorithm,
we introduce the notion of \emph{materialization graphs}, which
facilitates the analysis of the given algorithm.

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
\end{enumerate}
For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are two nodes,
we say that the node $v_1$ is the parent (node) of $v_2$ and the node $v_2$ is the child (node)
of $v_1$. Further, $\forall H,B_1,...,B_n\in V$, $\mathcal{G}$ satisfies the following conditions:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $H$ has a parent node or a child node;
  \item if $H$ has the in-degree of $0$, then $H$ is an original fact of $P$;
  \item $B_1,...,B_n\rightarrow H\in P^*$ is satisfied if $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$.
\end{enumerate}
\end{definition}

For some derived atom $H$, there may exist several rule instantiations where $H$
occurs as a head atom.
This also means that $H$ can be derived in different ways.
The condition in the definition above results in only one way of deriving $H$ being
described by a materialization graph.
Suppose $\mathcal{G}$ is a materialization graph such that the nodes
with in-degree 0 are the
original facts in $\textbf{I}$.
The size of $\mathcal{G}$, denoted by $|\mathcal{G}|$, is the number of nodes in $\mathcal{G}$.
The depth of $\mathcal{G}$, denoted by \texttt{depth}($\mathcal{G}$), is the maximal length of a path
in $\mathcal{G}$.
We next give an example of a materialization graph.\\

\begin{example}\label{exp:mg}
Consider a DHL($\circ$) ontology $\mathcal{O}_{ex_1}$ where the TBox is
$\{\exists R.A\sqsubseteq A\}$, the RBox is $\{S\circ R\sqsubseteq R\}$ and
the ABox is $\{A(b),R(a_1,b),S(a_i,a_{i-1})\}$ for $2\leq i\leq k$ and
$k$ is an integer greater than $2$.
The corresponding datalog program of this ontology is $P_{ex_1}=\langle R, \textbf{I}\rangle$
where $\textbf{I}$ contains all the assertions in the ABox
and $R$ contains the two rules `$R(x,y),A(y)\rightarrow A(x)$'
and `$S(x,y),R(y,z)\rightarrow R(x,z)$'.
The graph in Figure~\ref{fig:mg} is a materialization graph with respect to $P_{ex_1}$,
denoted by $\mathcal{G}_{ex_1}$.
The nodes with in-degree 0 are the original facts in $\textbf{I}$;
each of the other nodes corresponds to a ground instantiation of some rule.
For example, the node $A(a_k)$ corresponds to the ground rule instantiation
`$R(a_k,b),A(b)\rightarrow A(a_k)$'.
The size of this materialization graph is the number of nodes, that is $3k$.
The depth of $\mathcal{G}_{ex_1}$ is $k$.
\end{example}

@@

We say that a materialization graph $\mathcal{G}$ is a \emph{complete materialization graph}
when $\mathcal{G}$ contains all ground atoms in $T_R^{\omega}(\textbf{I})$.
The set of nodes in a complete materialization graph is actually the result
of materialization.
Thus, the procedure of materialization can be transformed into the construction of
a complete materialization graph.
In the remainder, we only consider complete materialization graphs and
do not distinguish the term from the
notion of `materialization graphs'.
It should also be noted that there may exist several materialization graphs for a datalog program.

\subsection{A Basic Parallel Algorithm}
\label{sec:alg-bsc}
